---
output:
  pdf_document: default
  html_document: default
---
# Output-Oriented Envelopment Model

Introduction
-----------------------------
In chapter 2 and 3 we focused on the input-oriented envelopment model. This chapter will focus on the output-oriented model.  People can sometimes be confused over when to use an input-oriented or an output-oriented model. Think of input-oriented as _input-reducing_ and output-oriented as _output increasing_. In an input-oriented model, the primary goal is to reduce the input while not sacrificing any output. In contrast, the primary goal for an output-oriented model is to increase the output while not increasing the usage of any input.

Think of selecting the model orientation by looking at a simple model of say, a set of restaurants where the input the input is labor and the output is customers served. The question than is what do the managers have more control over. If managers of restaurants have the opportunity to serve more customers, perhaps an output-oriented model is best. On the other hand, they have may have weak influence over the number of customers served, perhaps due to advertising being done at a regional level.  In this case, perhaps cost control is the primary goal which means that an input-oriented model should be applied to emphasize reducing labor.

In the case of constant returns to scale, the results will be consistent. Under other returns to scale assumptions, the orientation can have a significant impact on efficiency.  

The linear program for the output-oriented model is very similar to that of the input-oriented model.  The major change now is that rather finding the minimum level of input capable of producing the desired output, now the goal is to find the maximum output without reducing the input.  This expansion of output is denoted as $\phi$ and the resulting constant returns to scale model is given below.

$$
 \begin{aligned}
    \text{maximize  }   & \phi \\
    \text{subject to } & \sum_{j=1}^{n} x_{i,j}\lambda_j \leq  x_{i,k} \forall \; i\\
                       & \sum_{j=1}^{n} y_{r,j}\lambda_j \geq  \phi y_{r,k} \forall \; r\\
                       & \lambda_j \geq 0  \; \forall \; j
  \end{aligned}
$$

Just as we did in chapter 2 for the input-oriented model, we will transform it to standard form by moving all terms with variables to the left-hand side.

$$
 \begin{aligned}
    \text{maximize  }   & \phi \\
    \text{subject to } & \sum_{j=1}^{n} x_{i,j}\lambda_j \leq x_{i,k}  \forall \; i\\
                       & \sum_{j=1}^{n} y_{r,j}\lambda_j - \phi y_{r,k} \geq 0 \forall \; r\\
                       & \lambda_j \geq 0  \; \forall \; j
  \end{aligned}
$$

Now we will convert the problem input and output constraints from inequalities into equalities by explicitly defining slack variables.

$$
 \begin{aligned}
    \text{minimize  }   & \phi \\
    \text{subject to } & \sum_{j=1}^{n} x_{i,j}\lambda_j  + s^x_i =  x_{i,k} \forall \; i\\
                       & \sum_{j=1}^{n} y_{r,j}\lambda_j - \phi y_{r,k} - s^y_r = 0 \forall \; r\\
                       & \lambda_j , s^x_i, s^y_r \geq 0  \; \forall \; i,r,j
  \end{aligned}
$$

## Creating the Output-Oriented LP - The Algebraic Approach

As usual, we will start with loading a collection of packages.

```{r loading_packages, message=FALSE, warning=FALSE}
     # options: message=FALSE, warning=FALSE to turn off display

library(pander)           # Nicely formatted tables
library(dplyr)            # For data structure manipulation
library(ROI)              # R Optimization Interface package
library(ROI.plugin.glpk)  # Connection to glpk as solver
library(ompr)             # Optimization Modeling using R
library(ompr.roi)         # Connective tissue

```


```{r Creating_data_structures}

  XBaker1 <- matrix(c(150, 400, 320, 520, 350, 320, .2, 0.7, 1.2, 2.0, 1.2, 0.7),
                  ncol=2,dimnames=list(LETTERS[1:6],c("x1", "x2")))  
  YBaker1 <- matrix(c(14000, 14000, 42000, 28000, 19000, 14000, 3500, 21000, 10500, 
                    42000, 25000, 15000),
                  ncol=2,dimnames=list(LETTERS[1:6],c("y1", "y2")))

  ND <- nrow(XBaker1); NX <- ncol(XBaker1); NY <- ncol(YBaker1); # Define data size

  xdata<-XBaker1
  ydata<-YBaker1
                 # Now we will create lists of names
  DMUnames <- list(c(LETTERS[1:ND]))               # DMU names: A, B, ...
  Xnames<- lapply(list(rep("X",NX)),paste0,1:NX)   # Input names: x1, ...
  Ynames<- lapply(list(rep("Y",NY)),paste0,1:NY)   # Output names: y1, ...
  Vnames<- lapply(list(rep("v",NX)),paste0,1:NX)   # Input weight names: v1, ...
  Unames<- lapply(list(rep("u",NY)),paste0,1:NY)   # Output weight names: u1, ...
  SXnames<- lapply(list(rep("sx",NX)),paste0,1:NX) # Input slack names: sx1, ...
  SYnames<- lapply(list(rep("sy",NY)),paste0,1:NY) # Output slack names: sy1, ...
  Lambdanames<- lapply(list(rep("L_",ND)),paste0,LETTERS[1:ND])

  results.efficiency <- matrix(rep(-1.0, ND), nrow=ND, ncol=1)
#  dimnames(results.efficiency)<-list(c(LETTERS[1:ND]),c("Eff"))
  dimnames(results.efficiency)<-c(DMUnames,"CCR-OO")  # Attach names
  
  results.lambda     <- matrix(rep(-1.0, ND^2), nrow=ND,ncol=ND)
  dimnames(results.lambda)<-c(DMUnames,Lambdanames)

  results.xslack     <- matrix(rep(-1.0, ND*NX), nrow=ND,ncol=NX) 
  dimnames(results.xslack)<-c(DMUnames,SXnames)

  results.yslack     <- matrix(rep(-1.0, ND*NY), nrow=ND,ncol=NY) 
  dimnames(results.yslack)<-c(DMUnames,SYnames)

```

Let's refresh our memory with the input-output data for the example.

```{r ch4_baker_data, echo=FALSE}
pander (cbind(XBaker1, YBaker1),round=6, caption="Baker's Nursing Home Data")
```

Now the formulation is similar to the input-oriented model that we created in chapter 2.  The three changes are that we will be using $\phi$ rather than $\theta$, a _max_ rather than _min_ objective function, and modifying the output-constraint(s) rather than the input-constraint(s).

```{r Baker_example}

RTS<-"CRS"
for (k in 1:ND) {
  
  result <- MIPModel() %>%
    add_variable(vlambda[j], j = 1:ND, type = "continuous", lb = 0) %>%
    add_variable(vphi, type = "continuous") %>%
    set_objective(vphi, "max") %>%
    add_constraint(sum_expr(vlambda[j] * xdata[j,i], j = 1:ND) 
                   <=  xdata[k,i], i = 1:NX) %>%
    add_constraint(sum_expr(vlambda[j] * ydata[j,r], j = 1:ND) 
                   >= vphi * ydata[k,r], r = 1:NY) 

    if (RTS=="VRS") {result <- add_constraint(result, 
                 sum_expr(vlambda[j],j = 1:ND) == 1) }  #Returns to Scale

result <-   solve_model(result, with_ROI(solver = "glpk")) 
    
    results.efficiency[k] <-  get_solution(result, vphi)       
    results.lambda[k,] <- t(as.matrix(as.numeric(
                     get_solution(result, vlambda[j])[,3] )))
}
pander (cbind(results.efficiency, results.lambda),round=6, caption="Results from Baker's Example")

```

In this case, the two input-two output, output-oriented CCR model gives us four efficient nursing homes, A, B, C, and D.    

## Future Work

* Extend slack maximization model.
* Extract the dual weights for the inputs and outputs.  Requires new or undocumented features in the _ompr_ package.
* Add other RTS options.  Contrast efficiency scores and targets under varying RTS and orientations.

