---
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, eval=FALSE}
library(bookdown); library(rmarkdown); rmarkdown::render("05-Multiplier.Rmd", "pdf_book")
```

# Multiplier Models

## Introduction

Every linear program has a related alter ego referred to as the dual. By
duality, the two models have the same optimal objective function values.
In DEA, the multiplier models are simply the dual of the envelopment
models. In terms of the matrix representations of DEA given in chapter
3, the A matrix is transposed, rows give way to columns and columns
become rows. The right hand side values appear in the objective function
while the previous objective function coefficients become right hand
side values. More could be said about this but this will be deferred for
people interested in duality or more algorithmic aspects of linear
programming.

Let's turn our attention instead to intuitively developing the DEA
multiplier model through a few steps.

## The Two Output, No Input Model

Assume that you, Alfred, are one of six restaurant managers of various
franchisees of the popular TexMex chain, Chiposter. Each restaurant
branch is operating under different conditions and are operated by your
colleagues: Barb, Chris, Don, Ed, and Fran. Each restaurant manager is
achieving outcomes (outputs.) At the end of the year, you want to make a
case for having done well.

```{r multiplier-rest-Y-data }
library(kableExtra)

 YRest1 <- matrix(c(25, 14, 42, 18, 19, 14, 3500, 2100, 1050, 
                    4200, 2500, 1500), ncol=2,
                  dimnames=list(c("Alfred", "Barb","Chris", 
                                  "Don", "Ed","Fran"),
                                c("y1", "y2")))

ND <- nrow(YRest1); NY <- ncol(YRest1); # Define data size

kbl (YRest1, caption="Restaurant Data", booktabs = T)  

```

You can select any way of weighting the two outputs with just two
conditions:

-   No weight can be negative
-   No one can received a weighted score above 100%

Given these two limitations, your goal is to make your "Alfred score" as
high as possible.

While it is possible to experiment with different weights, this can also
be framed as an optimization model. The goal or objective is to find the
best Alfred score, using weights ($u_1$, and $u_2$) subject to the
constraints that no one's scores is above 100%.

Let's take the restaurant data and formulate the model.

$$
\begin{split}
\begin{aligned}
    \text {max }   & 25 u_1 + 3500 u_2 \\
    \text{s.t.:  } & 25 u_1 + 3500 u_2 \leq 1.0\\
                   & 14 u_1 + 2100 u_2 \leq 1.0\\
                   & 42 u_1 + 1050 u_2 \leq 1.0\\
                   & 18 u_1 + 4200 u_2 \leq 1.0\\
                   & 19 u_1 + 2500 u_2 \leq 1.0\\
                   & 14 u_1 + 1500 u_2 \leq 1.0\\
                   & u_1, u_2\geq 0  \
  \end{aligned}
\end{split}
$$

Next let's implement this in `ompr` which may serve as a convenient
refresher as well.

```{r, message=FALSE}
library(ROI, quietly=TRUE)              
library(ROI.plugin.glpk, quietly=TRUE)  
library(ompr, quietly=TRUE)             
library(ompr.roi, quietly=TRUE)     
```

```{r}

result <- MIPModel () |>
  add_variable (u1, type="continuous", lb=0) |>
  add_variable (u2, type="continuous", lb=0) |>
  set_objective (25*u1+3500*u2, "max") |>
  add_constraint (25*u1+3500*u2 <= 1.0) |>
  add_constraint (14*u1+2100*u2 <= 1.0) |>
  add_constraint (42*u1+1050*u2 <= 1.0) |>
  add_constraint (28*u1+4200*u2 <= 1.0) |>
  add_constraint (19*u1+2500*u2 <= 1.0) |>
  add_constraint (14*u1+1500*u2 <= 1.0) |>
  solve_model(with_ROI(solver = "glpk"))

u1_result <- get_solution(result, u1) 
u2_result <- get_solution(result, u2) 
alfred_score <- 25 * u1_result + 3500 * u2_result
names(alfred_score)<-""

kbl (cbind(alfred_score, u1_result, u2_result), 
       caption="Alfred's Score and Results", booktabs = T)
```

We could repeat the process for each person, by simply changing the
objective function.

```{r}
result <- MIPModel () |>
  add_variable (u1, type="continuous", lb=0) |>
  add_variable (u2, type="continuous", lb=0) |>
  set_objective (14*u1+2100*u2, "max") |>
  add_constraint (25*u1+3500*u2 <= 1.0) |>
  add_constraint (14*u1+2100*u2 <= 1.0) |>
  add_constraint (42*u1+1050*u2 <= 1.0) |>
  add_constraint (28*u1+4200*u2 <= 1.0) |>
  add_constraint (19*u1+2500*u2 <= 1.0) |>
  add_constraint (14*u1+1500*u2 <= 1.0) |>
  solve_model(with_ROI(solver = "glpk"))

u1_result <- get_solution(result, u1) 
u2_result <- get_solution(result, u2) 
barb_score <- 14 * u1_result + 2100 * u2_result
names(barb_score)<-""

kbl (cbind(barb_score, u1_result, u2_result), 
       caption="Barb's Score and Weights", booktabs = T, digits = 6)
```

Let's do it for Chris.

```{r}
result <- MIPModel () |>
  add_variable (u1, type="continuous", lb=0) |>
  add_variable (u2, type="continuous", lb=0) |>
  set_objective (42*u1+1050*u2, "max") |>
  add_constraint (25*u1+3500*u2 <= 1.0) |>
  add_constraint (14*u1+2100*u2 <= 1.0) |>
  add_constraint (42*u1+1050*u2 <= 1.0) |>
  add_constraint (28*u1+4200*u2 <= 1.0) |>
  add_constraint (19*u1+2500*u2 <= 1.0) |>
  add_constraint (14*u1+1500*u2 <= 1.0) |>
  solve_model(with_ROI(solver = "glpk"))

u1_result <- get_solution(result, u1) 
u2_result <- get_solution(result, u2) 
chris_score <- 42 * u1_result + 1050 * u2_result
names(chris_score)<-""

kbl (cbind(chris_score, u1_result, u2_result), 
       caption="Chris's Score and Weights", booktabs = T, digits = 6)
```

This process of building the model for each person individually and hard
coding the data into the model makes it difficult to maintain,
generalize, and apply to other cases.

Let's now generalize this by building the model algebraically. Let's
define the data, $y_{r,j}$ to be the value of output *r* for manager
*j*. Therefore, in this application, $y_{1,1}=25$ and $y_{2,1}=3500$
reflects Alfred (*j=1*) has outputs *1* and *2* of 25 and 3500
respectively.

We can then rewrite the formulation to the following. The summation is
over the two outputs.

$$
\begin{split}
\begin{aligned}
    \text {max } & \sum_{r=1}^{2} u_r y_{r,1}\\
    \text{s.t.:  } &  \sum_{r=1}^{2} u_r y_{r,1} \leq 1.0 \quad \text{[Alfred]}\\
     &  \sum_{r=1}^{2} u_r y_{r,2} \leq 1.0  \quad \text{[Barb]}\\
     &  \sum_{r=1}^{2} u_r y_{r,3} \leq 1.0  \quad \text{[Chris]}\\
     &  \sum_{r=1}^{2} u_r y_{r,4} \leq 1.0  \quad \text{[Don]}\\
     &  \sum_{r=1}^{2} u_r y_{r,5} \leq 1.0  \quad \text{[Ed]}\\
     &  \sum_{r=1}^{2} u_r y_{r,6} \leq 1.0  \quad \text{[Fran]}\\
     &                  u_1, u_2\geq 0  
  \end{aligned}
\end{split}
(\#eq:Ch5NLP6PeopleHArdCode)
$$

Now we will extend this further to reflect that the constraint of being
less than or equal to 1.0 is the same for every manager. A key notation
to learn is the $\forall$ symbol which means to repeat for all possible
values of the index.

$$
\begin{split}
\begin{aligned}
    \text {max }    & \sum_{r=1}^{2} u_r y_{r,1}\\
    \text{s.t.:   } &  \sum_{r=1}^{2} u_r y_{r,j} 
                          \leq 1.0  \; \forall \; j\\
                    &  u_r \geq 0  \; \forall \; r
  \end{aligned}
\end{split}
$$

The $\forall\;j$ in the constraints indicates that the constraint should
be repeated for all possible values of j. In our case it would have a
constraint for $j=1$, (Alfred), then another constraint with $j=2$
(Barb), etc. The $\forall\;r$ in variable lower bounds is used to repeat
the non-negativity of each weighting variable (in this case, $u_1$ and
$u_2$).

The above formulation is nearly generalized. Let's replace the last
hard-coded items remaining. The first is that the formulation only
calculates the score for first manager. To generalize this, let's
replace the *1* in the objective function with *k*. This would allow us
to calculate the optimal score for any manager, *k*. Secondly, the
summations each assume that there are only two outputs. Let's replace
this with $N^Y$ to serve as a count of the number of outputs.

$$
\begin{split}
\begin{aligned}
    \text {max }    & \sum_{r=1}^{N^Y} u_r y_{r,k}\\
    \text{s.t.:   } &  \sum_{r=1}^{N^Y} u_r y_{r,j} 
                          \leq 1.0 \; \forall \; j\\
                    & u_r \geq 0  \; \forall \; r
  \end{aligned}
\end{split}
$$

This formulation is now a linear programming model that will find an
optimal score for manager *k* regardless of the number of outputs and
the number of other managers being considered. This approach could be
used in a variety of peer evaluation or self-evaluation applications
where each unit is given the opportunity to evaluate themselves relative
to their peers. This model will always give a person their highest
possible score. Note that this model does not incorporate the resources
used to achieve these outcomes. For that, we can use a traditional ratio
idea of output over input.

## The Ratio Model

Now, let's further extend this model to incorporate inputs. Not only are
the restaurant managers producing different outcomes, they are also
using different resources, say capital and labor. Let's denote capital
investment as $x_1$ and labor in full-time-equivalents as $x_2$.

We can frame the question of doing a two-input, two-output study in the
same way as we did for the two output example earlier. Instead of a
simple score, the manager can have a ratio of weighted outputs over
weighted inputs.

For the two-input, two-output case, the formulation is given below.

$$
\begin{split}
\begin{aligned}
    \text {max } & \frac{\sum_{r=1}^{2} u_r y_{r,k}} {\sum_{i=1}^{2} v_i x_{i,k} } \\
    \text{s.t.:   } & \frac{\sum_{r=1}^{2} u_r y_{r,k}} {\sum_{i=1}^{2} v_i x_{i,k} }
                          \leq 1 \; \forall \; j\\
                    & u_r, v_i\geq 0  \; \forall \; r,\; i
  \end{aligned}
\end{split}
$$

We have two sets of variables now. The first is $u_r$ which is the
weight on the *r*'th output. The second is $v_i$ which is the weight on
the *i*'th input.

For clarification, we make a small variation from traditional DEA
notation. Normally the number of inputs is *m* and the number of outputs
is *s*. To make the code more readable, I will use a convention of *NX*
instead of *m* to refer to the number of inputs (x's) and *NY* to be the
number of outputs (y's) instead of *s*. Also, *n* is used to denote the
number of Decision Making Units (DMUs) and therefore I'll use *ND* to
indicate that in the R code. In the mathematical formulations,
superscripts are used to differentiate the the different N's resulting
in $N^X$, $N^Y$, and $N^D$ for `NX`, `NY`, and `ND` or *m*, *s*, and *n*
respectively.

$$
\begin{split}
\begin{aligned}
    \text {max } & \frac{\sum_{r=1}^{N^Y} u_r y_{r,k}} {\sum_{i=1}^{N^X} v_i x_{i,k} } \\
    \text{s.t.: } & \frac{\sum_{r=1}^{N^Y} u_r y_{r,k}} {\sum_{i=1}^{N^X} v_i x_{i,k} }
                          \leq 1 \; \forall \; j\\
                  &   u_r, \;v_i\geq 0  \; \forall \; r,i
  \end{aligned}
\end{split}
$$

This isn't a linear program because we are dividing functions of
variables by functions of variables. We need to make a few
transformations. First, we clear the denominator of each of the
constraints and collect the terms on resulting in the following
formulation.

$$
\begin{split}
 \begin{aligned}
    \text {max }   & \frac{\sum_{r=1}^{N^Y} u_r y_{r,k}} {\sum_{i=1}^{N^X} v_i x_{i,k} } \\
    \text{s.t.:  } & \sum_{r=1}^{N^Y} u_r y_{r,k} - \sum_{i=1}^{N^X} v_i x_{i,k} 
                          \leq 0 \; \forall \; j\\
                   & u_r, \; v_i\geq 0  \; \forall \; r,i
  \end{aligned}
 \end{split}
$$

Alas, it is still a nonlinear objective function. Fortunately, we can
take advantage of the structure of the problem. There are an infinite
number of possible combinations of numerators and denominators that can
give the same ratio. For example, if a unit were to receive a 90%
efficiency score, this could be with a numerator of nine and a
denominator of 10 or a numerator of 90 and a denominator of 100. Each of
these pairs satisfies the constraints and is therefore equally valid.
One clever trick solves the non-linearity by setting the denominator to
be a constant. The next step is to select normalizing value for the
objective function. Let's set the denominator equal to one. In this
case, we simply add a constraint, $\sum_{i=1}^{N^X} v_i x_{i,k}$, to the
linear program and replace the denominator in the objective function to
one means that we can rewrite the objective function using only the
numerator.

$$
\begin{split}
 \begin{aligned}
    \text {max   } & \sum_{r=1}^{N^Y} u_r y_{r,k} \\
    \text{s.t.:  } & \sum_{i=1}^{N^X} v_i x_{i,k} = 1 \\
    & \sum_{r=1}^{N^Y} u_r y_{r,k} - \sum_{i=1}^{N^X} v_i x_{i,k} 
                          \leq 0 \; \forall \; j\\
                   &  u_r, \; v_i\geq 0  \; \forall \; r,i
  \end{aligned}
 \end{split}
$$

This linear program can then be implemented in an linear programming
system.

## Creating the LP - The Algebraic Approach

We will implement this using the `ompr` package again, as we did in
chapters 2 and 4.

We're going to use our data from earlier of grocery stores.

```{r, echo=FALSE}
x <- matrix(c(10, 20, 30, 50, 22, 29,
              75, 100, 220, 480, 290, 210), byrow=FALSE, ncol=2, 
            dimnames=list(LETTERS[1:6],c("x1","x2")))
y <- matrix(c(75,100,300,400, 280, 120),ncol=1,
            dimnames=list(LETTERS[1:6],"y"))

storenames<- c("Al\'s Pantry", "Bob\'s Mill", 
               "Trader Carrie\'s", "Dilbertson\'s",
               "Ed\'s Eggshop", "Flo\'s Farmacopia")
temp<-cbind(storenames,x,y)
colnames(temp)<-c("Store Name", "Employees (x1)", 
                  "Floor Space (x2)", "Sales (y)")
kbl (temp, booktabs=TRUE)
NX <- ncol(x); NY <- ncol(y); ND <- nrow(x)
store7names <- TRA::DEAnames(NX, NY, ND)

ND <- nrow(x); NX <- ncol(x); NY <- ncol(y); 
# Define data size

xdata      <- x [1:ND,]  # Call it xdata
dim(xdata) <- c(ND,NX)  # structure data correctly
ydata      <- y [1:ND,]
dim(ydata) <- c(ND,NY)

```

Remember the inputs are named as "x1" and"x2" to represent the staff
(full-time equivalent employees) and the store size ($m^2$)
respectively. The output is sales are named "y1". As usual, I'm copying
the source data into objects, `xdata` and `ydata`, for actual operation.

```{r Structure-Results}

mm1.efficiency <- matrix(rep(-1.0, ND), nrow=ND)
mm1.lambda     <- matrix(rep(-1.0, ND^2), nrow=ND)
mm1.vweight    <- matrix(rep(-1.0, ND*NX), nrow=ND) 
mm1.uweight    <- matrix(rep(-1.0, ND*NY), nrow=ND) 
mm1.xslack     <- matrix(rep(-1.0, ND*NX), nrow=ND) 
mm1.yslack     <- matrix(rep(-1.0, ND*NY), nrow=ND) 

mm1names <- TRA::DEAnames(NX, NY, ND)
```

We are now ready to do the analysis. In chapter 2 we used the
envelopment model to examine this case. Now we will use the multiplier
model.

```{r eval=TRUE}
for (k in 1:ND) {

  result <- MIPModel() |>
  add_variable(vweight[i], i = 1:NX, type = "continuous", lb = 0)   |>
  add_variable(uweight[r], r = 1:NY, type = "continuous", lb = 0)   |>
  set_objective(sum_expr(uweight[r] * ydata[k,r], r = 1:NY), "max") |>
  add_constraint(sum_expr(vweight[i] * xdata[k,i], i = 1:NX) == 1)  |>
  add_constraint((sum_expr(uweight[r] * ydata[j,r], r = 1:NY)-
                    sum_expr(vweight[i] * xdata[j,i], i = 1:NX)) 
                 <= 0, j = 1:ND)
  result

  result <- solve_model(result, with_ROI(solver = "glpk", 
                                         verbose = FALSE))
  mm1.efficiency[k] <- objective_value (result) 

  # Get the output weights
  tempvweight <- get_solution(result, vweight[i])
  mm1.vweight[k,] <- tempvweight[,3]

  # Get the input weights
  tempuweight <- get_solution(result, uweight[i])
  mm1.uweight[k,] <- tempuweight[,3]
  templambda <- as.matrix(get_row_duals(result))
  mm1.lambda[k,] <- templambda [-1]
  # Drops the first dual value since that 
  #    constraint since is for setting denom=1
 }

mm1.combined <- cbind (mm1.efficiency, 
                       mm1.vweight, mm1.uweight,
                       mm1.lambda)
```

```{r}
kbl (mm1.combined , booktabs=T, escape=F, digits=4,
     col.names=c("$\\theta^{CRS}$", mm1names$VnamesLX,
                 mm1names$UnamesLX,
                 mm1names$LambdanamesbyletterLX),
     caption =c("Multiplier Model Results from 7 Store Example"))  |>
  kable_styling (latex_options = c("hold_position"))
```

The weights from the multiplier model often have multiple alternative
solutions-in other words, different values for decision variables that
give the same objective function value. In optimization, this situation
is called multiple optima. In DEA this means that the weights may be
different depending upon the particular settings used for solving the
linear program. The result is that you should be careful in over
interpreting the weights as well as in how you interpret results when
trying to reproduce analyses.

### Envelopment Multiplier Model Duality Relationship

Duality is a fundamental characteristic of linear programming. Every
linear program has a dual program that is a bit of mirror image of the
original. The multiplier model and the envelopment models are
fundamentally connected by duality. The multiplier model of DEA is
considered the dual of the envelopment model and the reverse is also
true.

Consider the basic structure of an LP as discussed in Chapter 3.

$$
\begin{split}
 \begin{aligned}
    \text{min  }   & CX \\
    \text{s.t.:  } & AX \geq B\\ 
                       & X \geq 0  
  \end{aligned}
   \end{split}
$$

A linear program can be converted to its dual by: - Creating a variable
for every constraint in the original LP - Creating a constraint for
every variable in the original LP - Transposing the *A* matrix,
sometimes denoted as $A^{\tau}$ - Using the *B* vector as the objective
function coefficients - Using the *C* vector as the right hand side
values of the constraints. - Changing the objective function from a
*min* to a *max* - Reverse the directions of the constraint inequalities

$$
\begin{split}
 \begin{aligned}
    \text{max  }   & BY \\
    \text{s.t.:  } & A^{\tau} Y \leq C\\ 
                       & Y \geq 0  \\
  \end{aligned}
   \end{split}
$$

We won't dwell further on this transformation and will leave it to the
reader to explore further but this is is why we can extract dual values
of the envelopment model to get the multiplier weights from the
multiplier model. Also, it is why we can extract the envelopment's
$\lambda$ values from the multiplier model's results.

(Example of extracting lambdas, compare to ch 2 results.)

## Weight Restrictions in the Multiplier Model

Simple example, then use a package

## Variable Returns to Scale

Formulate and implement a model Implement model
