---
output:
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE, eval=FALSE}
library(bookdown); library(rmarkdown); rmarkdown::render("06-Cross.Rmd", "pdf_book")
```

# Cross-Efficiency

## The Ratio Model

Recall the explanation of DEA from a ratio model perspective. It can be thought of as analogous to each DMU doing a self-evaluation relative to their peers. They can pick any non-negative weights on inputs or outputs that they want as long as they no one looks better than perfect. The ratio or multiplier model essentially stops there (with the caveat that there are extensions such as weight restrictions, super-efficiency, and returns to scale as discussed in the previous chapter.)

The Cross-Efficiency model goes one step further and says let's retain the imputed score that each DMU gives to each of the other DMUs. Instead of discarding all this extra information, it starts with the premise that these scores also have value. In a DEA model, DMU *k*'s efficiency is at least as good or better than what any of its peers would have given it, otherwise DMU *k* would have used those same weights to get that score.

## The Linear Programs for DEA

On the other hand, what if it allowed for \index{Blending} blending of units. There are a few assumptions that we could make. Let's start by saying that we can compare any particular products by rescaling (up or down) any other product as well as any combination of units.

Similarly, $x_{i,j}$ is the amount of input *i* used by unit *j* by DMU *j* and $y_{r,j}$ is the amount of output *r* used by DMU *j*.

This can be easily expanded to the multiple input and multiple output case by defining $x_i,j$ to be the amount of the *i*'th input used by unit *j* and $y_{r,j}$ to be the amount of the *r*'th output produced by unit *j*. While *m* and *s* are commonly used in DEA for the number inputs and outputs, to make the code more readable and help avoid name space conflicts, I will use `NX` instead of *m* to refer to the number of inputs (x's) and `NY` to be the number of outputs (y's) instead of *s*. Also, *n* is used to denote the number of Decision Making Units (DMUs) and therefore I'll use $N^D$ for this mathematically and `ND` to indicate that in the R code.

We have two important sets of variables now. The first is $u_r$ which is the weight on the *r*'th output. The second is $v_i$ which is the weight on the *i*'th input.

The input-oriented multiplier model can be thought of as finding a weighting scheme for outputs over inputs that give you the best possible score while giving no one better than *1.0*. For the sake of brevity, I'll show the models based on just the IO-multiplier model but the same principals hold and in general cross-efficiency can be conducted on the other DEA models.

$$\begin{split}
 \begin{aligned}
    \text {max } & \frac{\sum_{r=1}^{N^Y} u_r y_{r,k}} {\sum_{i=1}^{N^X} v_i x_{i,k} } \\
    \text{s.t.:  } & \frac{\sum_{r=1}^{N^Y} u_r y_{r,j}} {\sum_{i=1}^{N^X} v_i x_{i,j} }
                          \leq 1 \forall \; j\\
                  & u_r, v_i\geq 0  \; \forall \; r,i
  \end{aligned}
\end{split}
(\#eq:Ch6LPCCRIOM-Ratio-1a)$$

Just as in Chapter 5, this isn't yet a linear program because we are dividing functions of variables by functions of variables. We need to make a few transformations. First, we clear the denominator of each of the constraints resulting in the following formulation and then normalize the results to a single solution by setting the denominator of the objective function equal to one in a new constraint. See the previous chapter for more details on this transformation.

$$\begin{split}
 \begin{aligned}
    \text {max } & \sum_{r=1}^{N^Y} u_r y_{r,k} \\
    \text{s.t.:  } & \sum_{i=1}^{N^X} v_i x_{i,k} \\
    & \sum_{r=1}^{N^Y} u_r y_{r,j} - \sum_{i=1}^{N^X} v_i x_{i,j} 
                          \leq 0 \; \forall \; j\\
                  &  u_r, v_i\geq 0  \; \forall \; r,i
  \end{aligned}
\end{split}
(\#eq:Ch6LPCCRIOM-NoSlack)  $$

Cross-efficiency is then typically calculated as the average of each of these peer scores.

While implementation requires the linear program model, for understanding cross-efficiency, it is useful to go back to the ratio model. Now we will redefine $\theta_{k,j}$ as the cross-evaluation score of DMU *j* on the basis of the weights that are optimal for DMU *k*. The value, $\theta_{k,k}$ would be the regular DEA efficiency score.

$$\begin{split}
\begin{aligned}
    \text {max } & \theta_{k,k}=\frac{\sum_{r=1}^{N^Y} u_{r,k} y_{r,k}} {\sum_{i=1}^{N^X} v_i,k x_{i,k} } \\
    \text{s.t.:  } & \theta_{k,j}=\frac{\sum_{r=1}^{N^Y} u_{r,k} y_{r,j}} {\sum_{i=1}^{N^X} v_{i,k} x_{i,j} }
                          \leq 1 \; \forall \; j\\
            & u_{r,k}, v_{i,k}\geq 0  \; \forall \; r,i
  \end{aligned}
\end{split}
(\#eq:Ch6LPCCRIO-Cross1)$$

The cross-efficiency for DMU *j* is now simply the average what all DMUs gave *j* on their own scoring model. We could term this $CE_j$ or $\bar{\theta}_{\cdot j}$.

$$\begin{split}
 \begin{aligned}
    CE_j= \bar{\theta}_{\cdot j}=\frac {\sum_{k=1}^{N^D}\theta_{k,j}} {N^D}
        = \frac {\sum_{k=1}^{N^D} {
                        \frac {\sum_{r=1}^{N^Y} u_{r,k} y_{r,j} } 
                       {\sum_{i=1}^{N^X} v_{i,k} x_{i,j} } }    }
                {N^D}
  \end{aligned}
\end{split}
(\#eq:Ch6Calc-Cross)$$

## Implementing Cross-Efficiency

Let's return to our two-input, single-output, six grocery store example.
We will implement this using the `ompr` package again.

```{r, eval=TRUE}
library(kableExtra)           
library(dplyr)            
library(ROI)         
library(ROI.plugin.glpk)  
library(ompr)             
library(ompr.roi) 
```


```{r, echo=FALSE}
x <- matrix(c(10, 20, 30, 50, 22, 29,
              75, 100, 220, 480, 290, 210), byrow=FALSE, ncol=2, 
            dimnames=list(LETTERS[1:6],c("x1","x2")))
y <- matrix(c(75,100,300,400, 280, 120),ncol=1,
            dimnames=list(LETTERS[1:6],"y"))

storenames<- c("Al\'s Pantry", "Bob\'s Mill", 
               "Trader Carrie\'s", "Dilbertson\'s",
               "Ed\'s Eggshop", "Flo\'s Farmacopia")
temp<-cbind(storenames,x,y)
colnames(temp)<-c("Store Name", "Employees \n (x1)", 
                  "Floor Space (x2)", "Sales (y)")

kbl (temp, booktabs=TRUE, escape=FALSE)

ND <- nrow(x); NX <- ncol(x); NY <- ncol(y) # Define data size
xdata       <- x [1:ND,]   # Call it xdata
dim (xdata) <- c (ND,NX)   # structure data correctly
ydata       <- y [1:ND,]
dim (ydata) <- c (ND,NY)
```

Let's take another step from the previous DEA implementations by creating a function to do the input-oriented multiplier model. We'll simply pass the input and output data to model.

```{r CE-Function, eval=TRUE}

multiplierIO <- function (x,y) 
  { 
  ND <- nrow(x); NX <- ncol(x); NY <- ncol(y); # Define data size
  
  results.efficiency <- matrix(rep(-1.0, ND), nrow=ND, ncol=1)
  results.lambda     <- matrix(rep(-1.0, ND^2), nrow=ND,ncol=ND)
  results.vweight    <- matrix(rep(-1.0, ND*NX), nrow=ND,ncol=NX) 
  results.uweight    <- matrix(rep(-1.0, ND*NY), nrow=ND,ncol=NY) 

    for (k in 1:ND) {

    result <- MIPModel()                                               |>
    add_variable(Vvweight[i], i = 1:NX, type = "continuous", lb = 0)   |>
    add_variable(Vuweight[r], r = 1:NY, type = "continuous", lb = 0)   |>
    set_objective(sum_expr(Vuweight[r] * y[k,r], r = 1:NY), "max")     |>
    add_constraint(sum_expr(Vvweight[i] * x[k,i], i = 1:NX) == 1)      |>
    add_constraint((sum_expr(Vuweight[r] * y[j,r], r = 1:NY) -
                    sum_expr(Vvweight[i] * x[j,i], i = 1:NX)) 
                  <= 0, j = 1:ND)
    result

    result <- solve_model(result, with_ROI(solver = "glpk", 
                                           verbose = FALSE))
    results.efficiency[k] <- objective_value (result) 

    # Get the weights - Output weights
    tempvweight <- get_solution(result, Vvweight[i])
    results.vweight[k,] <- tempvweight[,3]

   # Get the weights- Input weights
    tempuweight <- get_solution(result, Vuweight[i])
    results.uweight[k,] <- tempuweight[,3]

  } # End of for k loop
  
  resultlist <- list(efficiency=results.efficiency, 
                     vweight=results.vweight, 
                     uweight=results.uweight)
 
  return(resultlist)  

} # End of function  
```

This function will work for any DEA model with one or more inputs and one or more outputs and non-negative data. In a production-ready function, we would want to add tests for data problems, allow parameters to select different models, and perhaps return more information but the basic idea holds.

Now we can simply call the function, `multiplierIO` and store the results.

```{r}
resfunc <- multiplierIO(x,y)

DMUnames <- list(c(LETTERS[1:ND]))
Vnames<- lapply(list(rep("v",NX)),paste0,1:NX) # Input weight names: v1, ...
Unames<- lapply(list(rep("u",NY)),paste0,1:NY) # Output weight names: u1, ...

dimnames(resfunc$efficiency) <- c(DMUnames,"CCR-IO")
dimnames(resfunc$vweight)    <- c(DMUnames,Vnames)
dimnames(resfunc$uweight)    <- c(DMUnames,Unames)
```

```{r, echo=FALSE}
kbl (cbind(resfunc$efficiency,resfunc$vweight,resfunc$uweight), 
        caption="Efficiency Scores and Weights", booktabs = T) |>
    kable_styling(latex_options = c("HOLD_position"))
```

You can think of the cross-efficiency by going back to revisit the original ratio model of the multiplier model. Let's extend the definition of the input and output weights to reflect for which unit the weights were determined. For example, $u_{r,k}$ is then the weight on output *r* when analyzed from the perspective of DMU *k* and a similar interpretation applies to input weight.

$$\begin{split}
\begin{aligned}
    \text {max } & \theta_{k,k}=\frac{\sum_{r=1}^{N^Y} u_{r,k} y_{r,k}} {\sum_{i=1}^{N^X} v_i,k x_{i,k} } \\
    \text{s.t.:  } & \theta_{j,k}=\frac{\sum_{r=1}^{N^Y} u_{r,k} y_{r,j}} {\sum_{i=1}^{N^X} v_{i,k} x_{i,j} }
                          \leq 1 \; \forall \; j\\
            & u_{r,k}, v_{i,k}\geq 0  \; \forall \; r,i
  \end{aligned}
\end{split}
(\#eq:Ch6LPCCRIO-Cross1)$$

Now the cross-efficiency for a unit j can be calculated as the following. I refer to the value $\theta_{k,j}$ as the cross-evaluation score for unit *j* from the perspective of the evaluation of unit *k*.

The cross-efficiency score for unit *j* is simply the average of all of the cross-evaluation scores given to *j* by the other units when they do their analysis.

$$\begin{split}
 \begin{aligned}
    CE_j= \frac {\sum_{k=1}^{N^D}\theta_{k,j}} {N^D}
        = \frac {\sum_{k=1}^{N^D} {
                        \frac {\sum_{r=1}^{N^Y} u_{r,k} y_{r,j} } 
                       {\sum_{i=1}^{N^X} v_{i,k} x_{i,j} } }    }
                {N^D}
  \end{aligned}
\end{split}
(\#eq:Ch6Calc-Cross)$$

```{r}
results.crosseval  <- matrix(rep(-1.0, ND*ND), nrow=ND,ncol=ND) 
results.crosseff   <- matrix(rep(-1.0, ND)) 

for (k in 1:ND) {
   for (j in 1:ND) {results.crosseval[k,j] <- 
     sum(resfunc$uweight[k,]*y[j,])/sum(resfunc$vweight[k,]*x[j,])
                   }
}
for (k in 1:ND) {
  results.crosseff[k]<-mean(results.crosseval[,k])
}
dimnames(results.crosseval) <- c(DMUnames,DMUnames)
dimnames(results.crosseff) <- c(DMUnames)
```

```{r, echo=FALSE}
colnames(results.crosseff)<-c("Cross-Efficiency")
kbl (rbind(results.crosseval,t(results.crosseff)), 
     booktabs = T, digits=5, escape=F,
     caption="Cross-Evaluation Matrix and Cross-Efficiency Scores from Direct Calculation") |>
    row_spec (7, bold=TRUE) |>
  add_header_above(c("Weights\nFrom"=1,"Peer Evaluation of:"=6)) |>
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

Notice that the values on the diagonal, $\theta_{k,k}$ are the same as the original efficiency scores.

We are going to be doing the cross-efficiency calculations often so let's convert this to a function.

```{r}
calccrosseff <- function (x,y,vmatrix,umatrix) {

  ND <- nrow(x); NX <- ncol(x); NY <- ncol(y); # Define data size
  
  DMUnames <- rownames(x)
  
  if (is.null(DMUnames)) {DMUnames<-list(c(LETTERS[1:ND]))}  
  
  crosseval  <- matrix(rep(-1.0, ND*ND), nrow=ND, ncol=ND,
                       dimnames=list(DMUnames,DMUnames))
  crosseff   <- matrix(rep(-1.0, ND), nrow=ND, 
                       dimnames=list(DMUnames, "Cross-Eff") )

  for (k in 1:ND) {
     for (j in 1:ND) {crosseval[k,j]<-
       sum(umatrix[k,]*y[j,]) / sum(vmatrix[k,]*x[j,])
                   }
   }
  
  crosseff <- colMeans (crosseval)

    resultlist <- list(crosseval=crosseval, crosseff=crosseff)
 
  return(resultlist)
}
```

```{r, eval=TRUE, echo=FALSE}
rescross <- calccrosseff (x, y, resfunc$vweight, resfunc$uweight)

kbl (rbind(rescross$crosseval,t(rescross$crosseff)), 
     booktabs = T, digits=5,
     caption="Cross-Evaluation Matrix and Cross-Efficiency Scores from Our Function") |>
    row_spec (7, bold=TRUE) |>
  add_header_above(c("Weights\nFrom"=1,"Peer Evaluation of:"=6)) |>
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

## Dealing with Cross-Efficiency's Multiple Optima

\index{Multiple optima|(} As discussed in the chapter on the Multiplier model, DEA often has multiple optima, particularly for efficient DMUs. While this doesn't affect the radial efficiency scores in either the envelopment or the multiplier model, it can have a major impact on the cross-evaluation scores and therefore the \index{Cross-efficiency} cross-efficiency scores.

To examine this issue, we need to invoke a secondary objective function again in the same manner as we did for the envelopment model.

A variety of mechanisms exist for dealing with the issue of multiple optima. We'll use an approach similar to the two-phase, slack maximization phase. In the first phase, we find the best radial efficiency score for DMU k, ($\theta_{k,k}$) and then pass that optimal value ($\theta_{k,k}^*$) to the second phase with a constraint,  $\theta_{k,k}=\theta_{k,k}^*$.

In the second phase, rather than maximizing slacks, we can try to either either minimize or maximize the collective population's score. Note that in this case, we can define the population input (or output) either including the unit being studied by summing over all of *j* or all of *j* excluding the unit studied, *k*.

$$\begin{split} \begin{aligned}     \text {min/max } & \frac{\sum_{r=1}^{N^Y} u_{r,k} \sum_{j=1}^{N^D} y_{r,j}} {\sum_{i=1}^{N^X} v_{i,k} \sum_{j=1}^{N^D}x_{i,j} } \\     \text{s.t.:  } & \theta_{j,k}=\frac{\sum_{r=1}^{N^Y} u_{r,k} y_{r,j}} {\sum_{i=1}^{N^X} v_{i,k} x_{i,j} }                           \leq 1 \; \forall \; j\\ & \theta_{k,k}=\frac{\sum_{r=1}^{N^Y} u_{r,k} y_{r,k}} {\sum_{i=1}^{N^X} v_{i,k} x_{i,k} } = \theta^*_{k,k}\\             & u_{r,k}, v_{i,k}\geq 0  \; \forall \; r,i   \end{aligned} \end{split} (\#eq:Ch6LPCCRIO-2Phase)$$

This ratio model can be linearized in the same way as the original ratio DEA model.

$$\begin{split} \begin{aligned}     \text {min/max } & \sum_{r=1}^{N^Y} u_{r,k} \sum_{j=1}^{N^D} y_{r,j}\\         \text{s.t.:  } & \sum_{i=1}^{N^X} v_{i,k} \sum_{j=1}^{N^D}x_{i,j} = 1 \\ & \sum_{r=1}^{N^Y} u_{r,k} y_{r,j} - \sum_{i=1}^{N^X} v_{i,k} x_{i,j}                            \leq 0 \; \forall \; j\\ & \sum_{r=1}^{N^Y} u_{r,k} y_{r,k}-  \theta^*_{k,k} \cdot\sum_{i=1}^{N^X} v_{i,k} x_{i,k} = 0 \\             & u_{r,k}, v_{i,k}\geq 0  \; \forall \; r,i   \end{aligned} \end{split} (\#eq:Ch6LPCCRIO-2Phase)$$

Let's implement both phases as a function again. I'll name the function `multiplerIO2PBen`. The first part is obvious as it is a multiplier model and input-oriented. The `2P` indicates it is a two-phase model. The last part, `Ben` indicates it is a *benevelent* model which means that it in the second phase, each DMU is trying to make the entire population look as it good as it can my maximizing the population efficiency.

```{r Cross-Eval-Sec, eval=TRUE}
# Implements secondary objective function to resolve multiple optima

multiplierIO2PBen <- function (x, y) 
  { 
  ND <- nrow(x); NX <- ncol(x); NY <- ncol(y); # Define data size
  
  results.efficiency <- matrix(rep(-1.0, ND), nrow=ND, ncol=1)
  results.lambda     <- matrix(rep(-1.0, ND^2), nrow=ND,ncol=ND)
  results.vweight    <- matrix(rep(-1.0, ND*NX), nrow=ND,ncol=NX) 
  results.uweight    <- matrix(rep(-1.0, ND*NY), nrow=ND,ncol=NY) 

    for (k in 1:ND) {
      res1P <- MIPModel()                                              |>
      add_variable(Vvweight[i], i = 1:NX, type = "continuous", lb = 0)  |>
      add_variable(Vuweight[r], r = 1:NY, type = "continuous", lb = 0)  |>
      set_objective(sum_expr(Vuweight[r] * y[k,r], r = 1:NY), "max")    |>
      add_constraint(sum_expr(Vvweight[i] * x[k,i], i = 1:NX) == 1)     |>
      add_constraint((sum_expr(Vuweight[r] * y[j,r], r = 1:NY)-
                    sum_expr(Vvweight[i] * x[j,i], i = 1:NX)) 
                  <= 0, j = 1:ND)                                       |>
      solve_model(with_ROI(solver = "glpk", verbose = FALSE))

      theta_k_k <- objective_value(res1P) # For constraining Phase 2
      
      res2P <- MIPModel()                                                    |>
      add_variable(Vvweight[i], i = 1:NX, type = "continuous", lb = 0)       |>
      add_variable(Vuweight[r], r = 1:NY, type = "continuous", lb = 0)       |>
      set_objective(sum_expr(Vuweight[r] * sum(y[j,r]), r = 1:NY, j = 1:ND), 
                    "max")                                                   |>
          # Modified objective function for Phase 2
      add_constraint(sum_expr(Vvweight[i] * sum(x[j,i]), i = 1:NX, 
                              j = 1:ND) == 1)                                |>
      add_constraint((sum_expr(Vuweight[r] * y[j,r], r = 1:NY)-
                    sum_expr(Vvweight[i] * x[j,i], i = 1:NX)) <= 0, j = 1:ND)|>
          # Following constraint limits k's efficiency to that of Phase 2
      add_constraint(sum_expr(Vuweight[r] * sum(y[k,r]), r = 1:NY)-
                    sum_expr( Vvweight[i] * theta_k_k *sum(x[k,i]), i = 1:NX) 
                    == 0)                                                    |>

      solve_model(with_ROI(solver = "glpk", verbose = FALSE))
      
    results.efficiency[k] <- objective_value (res1P) # From Phase 1

    # Get the weights - Output weights
    tempvweight <- get_solution(res2P, Vvweight[i])
    results.vweight[k,] <- tempvweight[,3]

   # Get the weights- Input weights
    tempuweight <- get_solution(res2P, Vuweight[i])
    results.uweight[k,] <- tempuweight[,3]

  } # End of for k loop
  
  resultlist <- list(efficiency=results.efficiency, 
                     vweight=results.vweight, 
                     uweight=results.uweight)
 
  return(resultlist)  

} # End of function  
```

Let's call the benevolent cross-efficiency model and calculate the cross-efficiencies.

```{r}
resfunc2P <- multiplierIO2PBen(x,y)

dimnames(resfunc2P$efficiency) <- c(DMUnames,"CCR-IO")
dimnames(resfunc2P$vweight)    <- c(DMUnames,Vnames)
dimnames(resfunc2P$uweight)    <- c(DMUnames,Unames)

rescross2P <- calccrosseff (x, y, resfunc2P$vweight, resfunc2P$uweight)

```


```{r, eval=TRUE, echo=FALSE}
kbl (rbind(rescross2P$crosseval,t(rescross2P$crosseff)), 
     booktabs = T, digits=5,
     caption="Cross-Evaluation Matrix and Cross-Efficiency Scores from Our Benevolent Model") |>
    row_spec (7, bold=TRUE) |>
  add_header_above(c("Weights\nFrom"=1,"Peer Evaluation of:"=6)) |>
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

In this example, the cross-efficiency scores are pretty similar to the original efficiency scores.  *D* has a modest decline while *A* and *F* have very similar values. On the other hand *E* which drops from being efficient to only about 90% efficient.  This is all because two of their peers *B* and *F* both give *E* a low grade of 70%. In contrast C, is considered efficient by all of its peers and retains its 1.0 score.

We can create a separate function for the *malevolent* case where each DMU is holding their efficiency constant but otherwise wants to make the population efficiency scores as low as possible. The only change is converting the secondary objective function from `max` to `min`. The code chunk for this function is in the RMarkdown for the book but for the sake of readability, I set the code chunk option to `echo=FALSE` to have it not displayed.  

```{r, echo=FALSE}
multiplierIO2PMal <- function (x, y) 
  { 
  ND <- nrow(x); NX <- ncol(x); NY <- ncol(y); # Define data size
  
  results.efficiency <- matrix(rep(-1.0, ND), nrow=ND, ncol=1)
  results.lambda     <- matrix(rep(-1.0, ND^2), nrow=ND,ncol=ND)
  results.vweight    <- matrix(rep(-1.0, ND*NX), nrow=ND,ncol=NX) 
  results.uweight    <- matrix(rep(-1.0, ND*NY), nrow=ND,ncol=NY) 

    for (k in 1:ND) {
      res1P <- MIPModel()                                               |>
      add_variable(Vvweight[i], i = 1:NX, type = "continuous", lb = 0)  |>
      add_variable(Vuweight[r], r = 1:NY, type = "continuous", lb = 0)  |>
      set_objective(sum_expr(Vuweight[r] * y[k,r], r = 1:NY), "max")    |>
      add_constraint(sum_expr(Vvweight[i] * x[k,i], i = 1:NX) == 1)     |>
      add_constraint((sum_expr(Vuweight[r] * y[j,r], r = 1:NY)-
                    sum_expr(Vvweight[i] * x[j,i], i = 1:NX)) 
                  <= 0, j = 1:ND)                                       |>
      solve_model(with_ROI(solver = "glpk", verbose = FALSE))

      theta_k_k <- objective_value(res1P) # For constraining Phase 2
      
      res2P <- MIPModel()                                                    |>
      add_variable(Vvweight[i], i = 1:NX, type = "continuous", lb = 0)       |>
      add_variable(Vuweight[r], r = 1:NY, type = "continuous", lb = 0)       |>     
      set_objective(sum_expr(Vuweight[r] * sum(y[j,r]), 
                                 r = 1:NY, j = 1:ND), "min")                 |>                               
      # Modified objective function for Phase 2 of the Malevolent model
      add_constraint(sum_expr(Vvweight[i] * sum(x[j,i]), i = 1:NX, 
                              j = 1:ND) == 1)                                |>
      add_constraint((sum_expr(Vuweight[r] * y[j,r], r = 1:NY)-
                    sum_expr(Vvweight[i] * x[j,i], i = 1:NX)) <= 0, j = 1:ND)|>
      add_constraint(sum_expr(Vuweight[r] * sum(y[k,r]), r = 1:NY)-
                    sum_expr( Vvweight[i] * theta_k_k *sum(x[k,i]), i = 1:NX) 
                    == 0)                                                    |>

      solve_model(with_ROI(solver = "glpk", verbose = FALSE))
      
    results.efficiency[k] <- objective_value (res1P) # From Phase 1

    # Get the weights - Output weights
    tempvweight <- get_solution(res2P, Vvweight[i])
    results.vweight[k,] <- tempvweight[,3]

   # Get the weights- Input weights
    tempuweight <- get_solution(res2P, Vuweight[i])
    results.uweight[k,] <- tempuweight[,3]

  } # End of for k loop
  
  resultlist <- list(efficiency=results.efficiency, 
                     vweight=results.vweight, 
                     uweight=results.uweight)
 
  return(resultlist)  

} # End of function  

```



```{r, echo=TRUE}
multiplierIO2Flex <- function (x, y, phase) 
  { 
  ND <- nrow(x); NX <- ncol(x); NY <- ncol(y); # Define data size
  
  results.efficiency <- matrix(rep(-1.0, ND), nrow=ND, ncol=1)
  results.lambda     <- matrix(rep(-1.0, ND^2), nrow=ND,ncol=ND)
  results.vweight    <- matrix(rep(-1.0, ND*NX), nrow=ND,ncol=NX) 
  results.uweight    <- matrix(rep(-1.0, ND*NY), nrow=ND,ncol=NY) 

    for (k in 1:ND) {
      res1P <- MIPModel()                                               |>
      add_variable(Vvweight[i], i = 1:NX, type = "continuous", lb = 0)  |>
      add_variable(Vuweight[r], r = 1:NY, type = "continuous", lb = 0)  |>
      set_objective(sum_expr(Vuweight[r] * y[k,r], r = 1:NY), "max")    |>
      add_constraint(sum_expr(Vvweight[i] * x[k,i], i = 1:NX) == 1)     |>
      add_constraint((sum_expr(Vuweight[r] * y[j,r], r = 1:NY)-
                    sum_expr(Vvweight[i] * x[j,i], i = 1:NX)) 
                  <= 0, j = 1:ND)                                       |>
      solve_model(with_ROI(solver = "glpk", verbose = FALSE))

      theta_k_k <- objective_value(res1P) # For constraining Phase 2
      
      mod2P <- MIPModel()                                                    |>
      add_variable(Vvweight[i], i = 1:NX, type = "continuous", lb = 0)       |>
      add_variable(Vuweight[r], r = 1:NY, type = "continuous", lb = 0)       |>

      add_constraint(sum_expr(Vvweight[i] * sum(x[j,i]), i = 1:NX, 
                              j = 1:ND) == 1)                                |>
      add_constraint((sum_expr(Vuweight[r] * y[j,r], r = 1:NY)-
                    sum_expr(Vvweight[i] * x[j,i], i = 1:NX)) <= 0, j = 1:ND)|>
      add_constraint(sum_expr(Vuweight[r] * sum(y[k,r]), r = 1:NY)-
                    sum_expr( Vvweight[i] * theta_k_k *sum(x[k,i]), i = 1:NX) 
                    == 0)                                                    
      
      if (phase=="mal") {  
          mod2P <- set_objective(mod2P, sum_expr(Vuweight[r] * sum(y[j,r]), 
                                 r = 1:NY, j = 1:ND), "min") }                                                
      if (phase=="ben") {  
          mod2P <- set_objective(mod2P, sum_expr(Vuweight[r] * sum(y[j,r]), 
                                 r = 1:NY, j = 1:ND), "max") }              
      # Modified objective function for Phase 2 of the Malevolent model

      res2P <- solve_model(mod2P, with_ROI(solver = "glpk", verbose = FALSE))
      
    results.efficiency[k] <- objective_value (res1P) # From Phase 1

    # Get the weights - Input weights
    tempvweight <- get_solution(res2P, Vvweight[i])
    results.vweight[k,] <- tempvweight[,3]

   # Get the weights- Output weights
    tempuweight <- get_solution(res2P, Vuweight[i])
    results.uweight[k,] <- tempuweight[,3]

  } # End of for k loop

  rescross   <- calccrosseff (x, y, results.vweight, results.uweight)  
  
  resultlist <- list(efficiency=results.efficiency, 
                     vweight=results.vweight, 
                     uweight=results.uweight,
                     crosseval=rescross$crosseval,
                     crosseff=rescross$crosseff)
 
  return(resultlist)  

} # End of function  

```

```{r}
resfunc2PMal <- multiplierIO2PMal(x,y)

dimnames(resfunc2PMal$efficiency) <- c(DMUnames,"CCR-IO")
dimnames(resfunc2PMal$vweight)    <- c(DMUnames,Vnames)
dimnames(resfunc2PMal$uweight)    <- c(DMUnames,Unames)

rescross2PMal <- calccrosseff (x, y, resfunc2PMal$vweight, resfunc2PMal$uweight)

```


```{r, eval=FALSE, echo=FALSE}
kbl (rbind(rescross2PMal$crosseval,t(rescross2PMal$crosseff)), 
     booktabs = T, digits=5,
     caption="Cross-Evaluation Matrix and Cross-Efficiency Scores from Our Malevolent Model") |>
    row_spec (7, bold=TRUE) |>
  add_header_above(c("Weights\nFrom"=1,"Peer Evaluation of:"=6)) |>
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

Nearly ever DMU has their cross-efficiency go down in the malevelent case, except for *B* which had a very small increase.  On the other hand, even *C* has a cross-eficiency below 1.0 because *E* uses a weighting scheme favors itself and berates *C*.  

We could create a more general function to handle a variety of other cases such as parameters to switch between malevolent and benevolent in the second phase,  output-orientation, alternative returns to scale, or weight restrictions. Essentially this would just be variations of how we use `if` logic on the `RTS` parameter back in Chapter 2. 

### Using another DEA Package

Let's explore the impact of the secondary objective function. The \index{MultiplierDEA} `MultiplierDEA` package has a cross-efficiency function but the `Mal_Ben` is a specialized version of the cross-efficiency function that implements the secondary goal of either malevolently trying to minimize other people's scores or benevolently trying to maximize other people's scores, based on the work of Doyle and Green.

This can be framed as a two-phase model in the same way as the envelopment model slack maximization model.

Let's compare these to the arbitrary cross-efficiency function which does neither. Note that this does not guarantee the existence of a unique optimal solution in terms of input and output weights, it does greatly decrease the potential.

#### Examining the Six Grocery Store Example

```{r}
library (MultiplierDEA)
resfunc    <- MultiplierDEA::CrossEfficiency (x, y, rts="crs", 
                               orientation="input")
resfuncmal <- MultiplierDEA::Mal_Ben (x, y, rts="crs", 
                       orientation="input", phase="mal")
resfuncben <- MultiplierDEA::Mal_Ben (x, y, rts="crs", 
                       orientation="input", phase="ben")
```

```{r, echo=FALSE}
rownames(resfunc$ce_ave)<-"Single-Phase"
rownames(resfuncmal$ce_ave)<-"Malevolent"
rownames(resfuncben$ce_ave)<-"Benevolent"

kbl (rbind(t(cbind(x,y)), 
           resfunc$ce_ave, resfuncmal$ce_ave, resfuncben$ce_ave),
     booktabs = T, digits=4, 
     caption = "Cross-Efficiency Data and Results for  Different Secondary Objectives" )  |>
     row_spec (4:6, bold=TRUE) |>
     pack_rows ("Inputs", 1, 2) |>
#     pack_rows ("Output", 3) |>
#     pack_rows ("Cross-Efficiency Results", 6, 7) |>
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

The cross-efficiency results are very different numerically but the relative ranking is similar. In this case, nursing home A stays the top scoring unit in all three cross-efficiency models but is joined by D in the Benevolent model. Also, G is the lowest ranked in all three models.

#### Revisiting Takamura and Tone's 2003 about Relocating Japan's Capital

Next, let's review the implementation of multiple optima in Auro's multiplier model. To test it, we'll load data from Takamura and Tone's 2003 paper on sites for relocating the capital of Japan.

```{r}

XTT03 <- matrix(c(1,1,1,1,1,1,1),ncol=1,
                dimnames=list(LETTERS[1:7],"x"))

YTT03 <- matrix(c(5,7,8,4,9,10,4,
                  10,10,7,8,4,2,7,
                  3,3,5,3,4,10,7),
                ncol=3, dimnames=list(LETTERS[1:7], 
                                      c("C1","C2","C3")))

ceTT03    <- CrossEfficiency (XTT03, YTT03, rts="crs", 
                               orientation="input")
ceTT03mal <- Mal_Ben (XTT03, YTT03, rts="crs", 
                       orientation="input", phase="mal")
ceTT03ben <- Mal_Ben (XTT03, YTT03, rts="crs", 
                       orientation="input", phase="ben")

ceTT03mal2 <- multiplierIO2Flex (XTT03, YTT03,"mal")
ceTT03ben2 <- multiplierIO2Flex (XTT03, YTT03,"ben")

```

```{r, echo=FALSE}
rownames(ceTT03$ce_ave)<-"Single-Phase"
rownames(ceTT03mal$ce_ave)<-"Malevolent"
rownames(ceTT03ben$ce_ave)<-"Benevolent"
#rownames(ceTT03mal2$crosseff)<-"Malevolent-My Function"
#rownames(ceTT03ben2$crosseff)<-"Benevolent-My Function"

kbl (rbind(as.data.frame(t(cbind(XTT03, YTT03)),digits=1),
           as.data.frame(rbind(ceTT03$ce_ave, ceTT03mal$ce_ave, 
                               ceTT03ben$ce_ave, 
                               ceTT03mal2$crosseff, ceTT03ben2$crosseff), digits=5)), 
           booktabs = T, digits=5,
        caption = "Data and Cross-Efficiency Results for Takamura and Tone, 2003") |>
     row_spec (5:7, bold=TRUE) |>
     pack_rows ("Inputs", 1, 1) |>
     pack_rows ("Outputs", 2, 4) |>
#     pack_rows ("Cross-Efficiency Results", 5, 9) |>
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

In this case, site location B, is highest ranked in all three cross-efficiency models.

Comments about cross-efficiency functions from `DEAmultiplier` package.

- Mal_and_Ben is not defined as a function.
- `CrossEfficiency` function should allow secondary objective function choice of "Malevolent" or "Benevolent"
- Results from `CrossEfficiency` do not match Takamura and Tone.
- Naming convention from results of analysis. Would be good if column names follow convention like my "helper file"-For example lambdas could use L\_ as a prefix.
- data validation. Ex. allow "Ben", "ben" "benevolent", "Benevolent". Also, RTS could do "CRS" or "crs"

#### Simple Classroom Example

Let's try the Spring case

```{r}
XSpringL <- matrix(c(1,1,1,1,1,1,1,1), ncol=1,
                   dimnames=list(LETTERS[1:8],"x"))

YSpringL <- matrix(c(77,10,88,13,6,35,75,3,
                     75,4,34,33,10,55,61,2,
                     80,60,2,44,27,33,50,90,
                     39,44,81,69,88,77,33,8),
                   ncol=4, dimnames=list(LETTERS[1:8],
                                         c("HW","MQ","Proj", "Exam")))
SpringLres <- CrossEfficiency(XSpringL, YSpringL)
```

```{r, echo=FALSE}
res <- rbind (SpringLres$ceva_matrix, SpringLres$ceva_max, 
              SpringLres$ce_ave)
resrows <-cbind (c("A", "B", "C", "D", "E", "F", "G", "H", 
         "Efficiency", "Cross-Efficiency"))
kbl (cbind(resrows, as.data.frame(res,digits=4)), 
     row.names=F, booktabs = T, digits=4, escape=F,
        caption = "Results from Class Grading Example")    |>
     row_spec (9:10, bold=TRUE) |>
     pack_rows ("Cross-Evaluation Matrix", 1, 8)           |>
     pack_rows ("Results", 9, 10) |>
         kable_styling(latex_options = c("HOLD_position", "scale_down"))
```


#### Revisiting Liang, Wu, Cook, and Zhu's DEA game cross-efficiency paper

```{r, eval=FALSE}
Xgame <- matrix(c(7,7,7,5,9,7,4,6,5,5,9,8,6,8,5), ncol=3, byrow=TRUE)

Ygame <- matrix(c(4,4,7,7,5,7,6,2,3,6), ncol=2, byrow=TRUE)

# gameresM<-MultiplierDEA::Mal_Ben (Xgame, Ygame, rts="crs", 
#                                 orientation="input", phase="mal")
#gameresB<-MultiplierDEA::Mal_Ben (Xgame, Ygame, rts="crs", 
#                                 orientation="input", phase="mal")
  
gameresM <- multiplierIO2Flex (Xgame,Ygame,"mal")
gameresB <- multiplierIO2Flex (Xgame,Ygame,"ben")

rbind(gameresM$crosseff,gameresMB$crosseff)
```


```{r, echo=FALSE}
res <- rbind (SpringLres$ceva_matrix, SpringLres$ceva_max, 
              SpringLres$ce_ave)
resrows <-cbind (c("A", "B", "C", "D", "E", "F", "G", "H", 
         "Efficiency", "Cross-Efficiency"))
kbl (cbind(resrows, as.data.frame(res,digits=4)), 
     row.names=F, booktabs = T, digits=4, escape=F,
        caption = "Results from Class Grading Example")    |>
     row_spec (9:10, bold=TRUE) |>
     pack_rows ("Cross-Evaluation Matrix", 1, 8)           |>
     pack_rows ("Results", 9, 10) |>
         kable_styling(latex_options = c("HOLD_position", "scale_down"))
```


\index{Multiple optima|)}

## Future Issues for Cross-Efficiency Chapter

Issues to consider in the future for this chapter or other chapters:

- Fixed Weighting nature of certain model dimensions (no-wait to talk about it in baseball chapter?)
- Check to see if Auro is using population with or without k in mal_ben code.
- Add reference to Doyle and Green.
